% =====================================================================
% Methods draft: Online verification — dynamic normative envelopes and
% the conditional P(IPV_i | IPV_j) criterion (with pseudocode).
% Requires (already in main.tex preamble): amsmath, algorithm, algorithmic.
% This file is the same content inserted into main.tex; kept standalone
% for review. Placeholders marked [TO SPECIFY] must be reconciled with
% the actual IPV estimator and calibration code.
% =====================================================================

\subsection{Online verification: dynamic envelopes and the conditional criterion}

\paragraph{Notation.}
We consider a dyadic interaction between an ego agent $i$ and its principal counterpart
$j$. At time $t$ we observe a trailing window $W_t=\{x(t'):\,t-\Delta\le t'\le t\}$ of
kinematic states. A causal estimator $g$ returns a \emph{dynamic} interaction preference
value (IPV) for each agent, $\hat\theta_i(t)=g(W_t;i)$ and $\hat\theta_j(t)=g(W_t;j)$, with
$\theta\in[-\pi/2,\pi/2]$ and $\theta>0$ denoting prosocial behaviour. Because $g$ reads only
past observations, $\hat\theta(t)$ is a strict-online signal, distinct from the full-window,
case-mean IPV used only as an offline calibration label.
\emph{[TO SPECIFY: window length $\Delta$; whether predicted-intention trajectories augment
$W_t$; reported convergence/bias of $\hat\theta(t)$ relative to the full-window estimate.]}

\paragraph{Interaction state.}
We summarise the situation by a state $s(t)=(r(t),h,\rho_i,\tau(t))$, where $r(t)$ is an
online risk bin from the hybrid proxy $r=\min(\widehat{\mathrm{APET}},\mathrm{TTC}_{\mathrm{close}})$,
$h$ is the coarse geometry (path category), $\rho_i\in\{\text{priority},\text{non-priority}\}$
is the map-derived role, and $\tau(t)\in[0,1]$ is the interaction progress from a causal
rolling-phase detector. The progress index $\tau$ is essential: it makes the norm a function
of how far the interaction has developed, so that the online signal is compared with a norm
estimated at the same stage rather than with a whole-episode summary.

\paragraph{Dynamic normative envelope.}
Offline, for each state cell $c=(r,h,\rho_i)$ and progress $\tau$, we estimate the
distribution of the dynamic IPV computed with the \emph{same} window $\Delta$ on human data,
not the full-window mean. The marginal envelope is the set of conditional quantiles
\begin{equation}
Q_\alpha(c,\tau)=\inf\{\theta:\hat F(\theta\mid c,\tau)\ge\alpha\},\qquad
\text{reported as } [q_{05},q_{50},q_{95}].
\end{equation}
Calibrating on the rolling estimator removes the rolling-versus-full-window mismatch, so an
online value is scored against a like-for-like reference. Sparse cells back off to a parent
level by hierarchical partial pooling.

\paragraph{Conditional criterion $P(\theta_i\mid\theta_j)$.}
Human interactions show event-level coupling: simultaneous strong competition is rare, so an
agent's socially reasonable preference depends on its counterpart's. We therefore verify the
ego against the \emph{conditional} norm. For state $c$ and progress $\tau$ we estimate
$\hat F(\theta_i\mid\theta_j,c,\tau)$ and define the reasonable band
\begin{equation}
B_{1-\eta}(\theta_j,c,\tau)=\bigl[\,Q_{\eta/2}(\theta_j,c,\tau),\;Q_{1-\eta/2}(\theta_j,c,\tau)\,\bigr],
\end{equation}
the central $1-\eta$ conditional interval of ego IPV given the counterpart's current IPV.
When the counterpart is competitive ($\theta_j\ll0$) the band shifts prosocial; an ego that
is simultaneously competitive then falls outside the band, flagging the configuration that
empirically precedes conflict.
\emph{[TO SPECIFY: conditional quantiles via binning on $\theta_j$ within each cell, or a
monotone quantile regression; chosen per cell support.]}

\paragraph{Deviation score.}
The signed, normalised nonconformity of the ego is
\begin{equation}
\delta_i(t)=\frac{\hat\theta_i(t)-Q_{1/2}\!\bigl(\hat\theta_j(t),c,\tau\bigr)}
{w\!\bigl(\hat\theta_j(t),c,\tau\bigr)},
\qquad
w=\tfrac{1}{2}\bigl(Q_{1-\eta/2}-Q_{\eta/2}\bigr),
\end{equation}
so $|\delta_i(t)|>1$ marks a value outside the reasonable band and the sign gives the
direction (competitive versus over-yielding). For a calibrated flag we wrap $\delta$ in a
per-cell (Mondrian) split-conformal threshold $\kappa_c$ estimated on held-out human data; a
flag at level $\beta$ then means $\delta$ exceeds the $1-\beta$ human quantile for that cell.
Because coverage degrades under source shift, $\kappa_c$ is calibrated source-aware and
reported with its empirical coverage rather than assumed nominal. The online score is
smoothed by an exponentially weighted moving average,
$\bar\delta(t)=\lambda\,\delta_i(t)+(1-\lambda)\,\bar\delta(t-1)$, and persistence is required
before escalation.

\paragraph{Verdict and planner interface.}
A verdict is emitted only after quality gates pass: sufficient cell support, admissible input
provenance (leakage contract), bounded estimator uncertainty, and temporal hysteresis. The
gated score maps to a graded output: a monitor record always; a one-sided soft cost
$c_{\mathrm{social}}=\max(0,\bar\delta-\theta_{\mathrm{warn}})\cdot\mathbf{1}[\text{norm prosocial}]$;
a warning above $\theta_{\mathrm{warn}}$; and a fallback candidate above $\theta_{\mathrm{fb}}$
under sustained high risk.
\emph{[TO SPECIFY: thresholds $\theta_{\mathrm{warn}},\theta_{\mathrm{fb}}$ set on the
calibration set; demonstrative values used in pilot analyses were $0.25$ and $0.85$.]}

% ---------------------------------------------------------------------
\begin{algorithm}[t]
\caption{Offline calibration of dynamic and conditional envelopes}
\label{alg:calib}
\begin{algorithmic}[1]
\REQUIRE human interactions $\mathcal{D}$; window $\Delta$; levels $\alpha,\eta,\beta$;
         min support $n_{\min}$
\ENSURE marginal $Q$, conditional bands $B$, per-cell thresholds $\kappa$
\STATE split $\mathcal{D}$ into estimation set $\mathcal{D}_e$ and calibration set $\mathcal{D}_k$
\STATE $\mathcal{T}\leftarrow\emptyset$
\FOR{each interaction in $\mathcal{D}_e$ and each time $t$}
    \STATE $\theta_i,\theta_j\leftarrow g(W_t;i),\,g(W_t;j)$ \COMMENT{causal dynamic IPV}
    \STATE $(r,h,\rho_i,\tau)\leftarrow \textsc{RecognizeState}(W_t)$
    \STATE append $\bigl((r,h,\rho_i),\tau,\theta_i,\theta_j\bigr)$ to $\mathcal{T}$
\ENDFOR
\FOR{each cell $c$ and progress bin $\tau$ in $\mathcal{T}$}
    \IF{$\mathrm{support}(c,\tau)<n_{\min}$}
        \STATE attach $c$ to $\textsc{Parent}(c)$ \COMMENT{hierarchical fallback}
    \ENDIF
    \STATE $Q_\alpha(c,\tau)\leftarrow$ empirical quantiles of $\theta_i$
    \STATE $B_{1-\eta}(\theta_j,c,\tau)\leftarrow$ conditional quantiles of $\theta_i$ given $\theta_j$
\ENDFOR
\STATE on $\mathcal{D}_k$: $\kappa_c\leftarrow (1-\beta)$ quantile of $|\delta|$ per cell \COMMENT{Mondrian split-conformal, source-aware}
\RETURN $Q,\,B,\,\kappa$
\end{algorithmic}
\end{algorithm}

% ---------------------------------------------------------------------
\begin{algorithm}[t]
\caption{Online social-compliance verification step at time $t$}
\label{alg:online}
\begin{algorithmic}[1]
\REQUIRE window $W_t$; calibrated $Q,B,\kappa$; gates; previous EWMA $\bar\delta_{t-1}$;
         smoothing $\lambda$; thresholds $\theta_{\mathrm{warn}},\theta_{\mathrm{fb}}$
\ENSURE verdict and planner-facing signal
\STATE $\theta_i,\theta_j\leftarrow g(W_t;i),\,g(W_t;j)$ \COMMENT{strict-online; no future info}
\STATE $(r,h,\rho_i,\tau)\leftarrow \textsc{RecognizeState}(W_t)$
\STATE $c\leftarrow(r,h,\rho_i)$;\quad \textbf{if} $\mathrm{support}(c,\tau)<n_{\min}$ \textbf{then} $c\leftarrow\textsc{Parent}(c)$
\STATE $m\leftarrow Q_{1/2}(\theta_j,c,\tau)$;\quad $w\leftarrow\tfrac12\bigl(Q_{1-\eta/2}-Q_{\eta/2}\bigr)(\theta_j,c,\tau)$
\STATE $\delta\leftarrow (\theta_i-m)/w$ \COMMENT{signed conditional nonconformity}
\STATE $\bar\delta\leftarrow \lambda\,\delta+(1-\lambda)\,\bar\delta_{t-1}$ \COMMENT{EWMA}
\IF{\textbf{not} \textsc{QualityGates}$(c,\text{provenance},\text{uncertainty})$}
    \RETURN $\textsc{Monitor}(\bar\delta)$ \COMMENT{insufficient support/quality: record only}
\ENDIF
\IF{$\bar\delta>\kappa_c$ \textbf{and} $r$ high \textbf{and} persistent}
    \STATE $\text{signal}\leftarrow\textsc{FallbackCandidate}$
\ELSIF{$\bar\delta>\theta_{\mathrm{warn}}$}
    \STATE $\text{signal}\leftarrow\textsc{Warning}$, with soft cost $\max(0,\bar\delta-\theta_{\mathrm{warn}})\cdot\mathbf{1}[\text{norm prosocial}]$
\ELSE
    \STATE $\text{signal}\leftarrow\textsc{Monitor}$
\ENDIF
\RETURN $(\bar\delta,\ \text{signal})$
\end{algorithmic}
\end{algorithm}

\noindent\textbf{Complexity and determinism.} Each online step is $O(1)$ given precomputed
envelope tables (state recognition, two table look-ups and a scalar update), so the verifier
runs within a planning cycle and is deterministic given $W_t$, supporting auditability.
